---
title: "Assessment EDA"
author: "Conor Tompkins"
date: "1/26/2021"
output:
  html_document:
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE)
```

The purpose of this document is to perform EDA on the `clean_assessments.csv` data to identify patterns in the data that will inform a model of sale price.

```{r}
library(tidyverse)
library(lubridate)

library(sf)
library(leaflet)

library(skimr)
library(hrbrthemes)
library(janitor)
library(scales)

library(here)

#here::i_am("scripts/eda")

options(scipen = 999, digits = 4)

theme_set(theme_ipsum())

assessment_data_path <- here("data/cleaned/big", "clean_assessment_data_geocoded.csv")

unified_geo_ids_path <- here("data/cleaned/big/unified_geo_ids/unified_geo_ids.shp")

assessments_valid <- read_csv(assessment_data_path) %>% 
  mutate(sale_month = factor(sale_month, levels = month.abb))

geo_ids <- st_read(unified_geo_ids_path)
```

### Get an idea of the data

```{r}
glimpse(assessments_valid)
```

```{r}
skim(assessments_valid)
```

### Sales Price

Inflation-adjusted sales price (`sale_price_adj`) is normally distributed on the `log10()` scale, as expected.
```{r}
assessments_valid %>% 
  ggplot(aes(sale_price_adj)) +
  geom_density() +
  scale_x_log10(labels = dollar)
```

Adjusting for inflation (`sale_price_adj` is 2020 dollars) removes a lot of the drift over time.
```{r}
assessments_valid %>%
  select(sale_year, sale_price, sale_price_adj) %>% 
  pivot_longer(cols = contains("sale_price")) %>% 
  ggplot(aes(sale_year, value, color = name)) +
  geom_smooth() +
  scale_y_continuous(labels = dollar)
```
House sales have plateaued since the 2008 mortgage crisis.
```{r}
sale_year_count <- assessments_valid %>% 
  count(sale_year)

sale_year_count %>% 
  mutate(year_flag = case_when(sale_year == 2021 ~ "2021",
                          sale_year < 2021 ~ "Before 2021")) %>% 
  ggplot(aes(sale_year, n)) +
  geom_smooth(data = sale_year_count %>% 
                filter(sale_year != 2021),
              se = F) +
  geom_point(aes(color = year_flag)) +
  labs(x = "Sale Year",
       y = "Sales",
       color = NULL)
```

```{r}
assessments_valid %>% 
  mutate(sale_month = fct_rev(sale_month)) %>% 
  add_count(sale_month) %>% 
  ggplot(aes(sale_month, sale_price_adj, group = sale_month, fill = n)) +
  geom_boxplot(outlier.alpha = 0,
               color = "grey") +
  scale_y_log10() +
  coord_cartesian(ylim = c(10^4, 10^6)) +
  scale_fill_viridis_c()
```

### Location

The location of the house (`geo_id`) has a big impact on `sale_price_adj`.
```{r}
geo_id_median_price <- assessments_valid %>% 
  group_by(geo_id) %>% 
  summarize(median_price = median(sale_price_adj))

pal <- colorNumeric(
  palette = "viridis",
  domain = geo_id_median_price$median_price)

geo_ids %>% 
  left_join(geo_id_median_price) %>% 
  leaflet() %>% 
  addProviderTiles(providers$Stamen.TonerLite,
                   options = providerTileOptions(noWrap = TRUE,
                                                 minZoom = 9, 
                                                 #maxZoom = 8
                   )) %>% 
  addPolygons(popup = ~ str_c(geo_id, " ", "median price: ", dollar(median_price), sep = ""),
              fillColor = ~pal(median_price),
              fillOpacity = .7,
              color = "black",
              weight = 3) %>% 
  addLegend("bottomright", pal = pal, values = ~median_price,
            title = "Median sales price",
            labFormat = labelFormat(prefix = "$"),
            opacity = 1)
```

```{r fig.height=16}
assessments_valid %>%
  add_count(geo_id) %>% 
  mutate(geo_id = fct_reorder(geo_id, sale_price_adj, .fun = median)) %>% 
  ggplot(aes(sale_price_adj, geo_id, fill = n)) +
  geom_boxplot(outlier.alpha = 0,
               color = "grey") +
  scale_x_log10(label = dollar) +
  scale_fill_viridis_c()
```

### House Characteristics

The age of the house at time of sale has a negative relationship with sale price, but is heteroskedastic. This indicates that some houses retain their value better over time (and/or that houses that don't retain value aren't bought and sold).
```{r}
assessments_valid %>% 
  ggplot(aes(house_age_at_sale, sale_price_adj)) +
  geom_density_2d_filled() +
  scale_y_log10(labels = dollar)
```

#### Lot Area

There is not an obvious relationship between `lot_area` and `sale_price_adj`.
```{r}
assessments_valid %>% 
  ggplot(aes(lot_area, sale_price_adj)) +
  geom_density_2d_filled() +
  scale_y_log10(labels = dollar) +
  coord_cartesian(xlim = c(0, 50000))
```

But, `lot_area` varies drastically across `geo_id`, so there might still be a useful feature to be engineered.
```{r fig.height=16}
assessments_valid %>% 
  mutate(geo_id = fct_reorder(geo_id, lot_area, .fun = median)) %>% 
  ggplot(aes(lot_area, geo_id)) +
  geom_boxplot(outlier.alpha = 0) +
  coord_cartesian(xlim = c(0, 40000))
```

`lot_area` also varies across `style_desc`.
```{r fig.height=14}
assessments_valid %>% 
  mutate(style_desc = fct_reorder(style_desc, lot_area, .fun = median)) %>% 
  ggplot(aes(lot_area, style_desc)) +
  geom_boxplot(outlier.alpha = 0) +
  coord_cartesian(xlim = c(0, 10^5))
```

#### Finished Living Area

`finished_living_area` will be a useful variable.
```{r}
assessments_valid %>% 
  ggplot(aes(finished_living_area, sale_price_adj)) +
  geom_density_2d_filled() +
  scale_y_log10(labels = dollar) +
  coord_cartesian(xlim = c(0, 5000))
```

`finished_living_area` also varies across `style_desc`.
```{r fig.height=14}
assessments_valid %>% 
  mutate(style_desc = fct_reorder(style_desc, finished_living_area, .fun = median)) %>% 
  ggplot(aes(finished_living_area, style_desc)) +
  geom_boxplot(outlier.alpha = 0) +
  coord_cartesian(xlim = c(0, 10^4))
```

`finished_living_area` also varies across `geo_id`, but not as drastically.
```{r fig.height=16}
assessments_valid %>% 
  mutate(geo_id = fct_reorder(geo_id, finished_living_area, .fun = median)) %>% 
  ggplot(aes(finished_living_area, geo_id)) +
  geom_boxplot(outlier.alpha = 0) +
  coord_cartesian(xlim = c(0, 10000))
```

#### Grade

`grade_desc` has a clear relationship with `sale_price_adj`.
```{r}
assessments_valid %>% 
  mutate(grade_desc = fct_reorder(grade_desc, sale_price_adj, median)) %>% 
  add_count(grade_desc) %>% 
  ggplot(aes(sale_price_adj, grade_desc, fill = n)) +
  geom_boxplot(outlier.alpha = 0,
               color = "grey") +
  scale_x_log10(labels = dollar) +
  scale_fill_viridis_c()
```

#### CDU

```{r}
assessments_valid %>% 
  mutate(cdu_desc = fct_reorder(cdu_desc, sale_price_adj, .fun = median)) %>% 
  add_count(cdu_desc) %>% 
  ggplot(aes(sale_price_adj, cdu_desc,  fill = n)) +
  geom_boxplot(outlier.alpha = 0,
               color = "grey") +
  scale_x_log10(label = dollar) +
  scale_fill_viridis_c()
```

#### Condition

`condition_desc` also has a relationship, but some of the levels can probably be collapsed.
```{r}
assessments_valid %>% 
  mutate(condition_desc = fct_reorder(condition_desc, sale_price_adj, median)) %>% 
  add_count(condition_desc) %>% 
  ggplot(aes(sale_price_adj, condition_desc, fill = n)) +
  geom_boxplot(outlier.alpha = 0,
               color = "grey") +
  scale_x_log10(labels = dollar) +
  scale_fill_viridis_c()
```

There are 4 main types of houses, with a lot of low-n types that can be collapsed.
```{r fig.height=12}
assessments_valid %>% 
  count(style_desc) %>% 
  mutate(style_desc = fct_reorder(style_desc, n, .fun = max)) %>% 
  ggplot(aes(n, style_desc)) +
  geom_col()
```

There is some time-series pattern in when different types of houses were created.
```{r}
assessments_valid %>% 
  add_count(style_desc) %>% 
  filter(n > 5000) %>% 
  mutate(style_desc = fct_reorder(style_desc, n, .fun = "max", .desc = T)) %>% 
  ggplot(aes(year_built, fill = style_desc)) +
  geom_histogram(binwdidth = 30) +
  guides(fill = FALSE) +
  facet_wrap(~style_desc, scales = "free_y")
```

#### Bathrooms

Most homes have between 1 and 2 full and half bathrooms.
```{r}
assessments_valid %>% 
  count(full_baths, half_baths) %>% 
  complete(full_baths = 0:12, half_baths = 0:9, fill = list(n = 0)) %>% 
  ggplot(aes(full_baths, half_baths, fill = n)) +
  geom_tile() +
  scale_fill_viridis_c() +
  scale_x_continuous(breaks = c(0:12),
                     expand = c(0,0)) +
  scale_y_continuous(breaks = c(0:12),
                     expand = c(0,0)) +
  coord_equal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

`fullbaths` and `sale_price_adj` are positively related.
```{r}
assessments_valid %>% 
  add_count(full_baths) %>% 
  ggplot(aes(sale_price_adj, full_baths, fill = n, group = full_baths)) +
  geom_boxplot(outlier.alpha = 0,
               color = "grey") +
  scale_x_log10(label = dollar) +
  scale_y_continuous(breaks = c(0:12)) +
  scale_fill_viridis_c()
```

There appear to be diminishing returns on the number of half bathrooms.
```{r}
assessments_valid %>% 
  add_count(half_baths) %>% 
  ggplot(aes(sale_price_adj, half_baths, fill = n, group = half_baths)) +
  geom_boxplot(outlier.alpha = 0,
               color = "grey") +
  scale_x_log10(label = dollar) +
  scale_y_continuous(breaks = c(0:9)) +
  scale_fill_viridis_c()
```

#### Heating and cooling

Need to split these out to heat_type and ac_flag.
```{r}
assessments_valid %>% 
  mutate(heating_cooling_desc = fct_reorder(heating_cooling_desc, sale_price_adj, .fun = median)) %>% 
  add_count(heating_cooling_desc) %>% 
  ggplot(aes(sale_price_adj, heating_cooling_desc, fill = n)) +
  geom_boxplot(outlier.alpha = 0,
               color = "grey") +
  scale_x_log10(label = dollar) +
  scale_fill_viridis_c()
```

Impute missing based on mode for `geo_id` and `style_desc`.
```{r}
assessments_valid %>% 
  mutate(heat_type = fct_reorder(heat_type, sale_price_adj, .fun = median)) %>% 
  add_count(heat_type) %>% 
  ggplot(aes(sale_price_adj, heat_type, fill = n)) +
  geom_boxplot(outlier.alpha = 0,
               color = "grey") +
  scale_x_log10(label = dollar) +
  scale_fill_viridis_c()
```

Impute missing based on mode for `geo_id` and `style_desc.`
```{r}
assessments_valid %>% 
  mutate(ac_flag = as.character(ac_flag)) %>% 
  mutate(ac_flag = fct_reorder(ac_flag, sale_price_adj, .fun = median)) %>% 
  add_count(ac_flag) %>% 
  ggplot(aes(sale_price_adj, ac_flag, fill = n)) +
  geom_boxplot(outlier.alpha = 0,
               color = "grey") +
  scale_x_log10(label = dollar) +
  scale_fill_viridis_c()
```

#### Exterior

Impute missing based on mode for `geo_id` and `style_desc.`
```{r}
assessments_valid %>% 
  mutate(extfinish_desc = fct_reorder(extfinish_desc, sale_price_adj, .fun = median)) %>% 
  add_count(extfinish_desc) %>% 
  ggplot(aes(sale_price_adj, extfinish_desc, fill = n)) +
  geom_boxplot(outlier.alpha = 0,
               color = "grey") +
  scale_x_log10(label = dollar) +
  scale_fill_viridis_c()
```

#### Roof

Impute missing based on mode for `geo_id` and `style_desc.`
```{r}
assessments_valid %>% 
  mutate(roof_desc = fct_reorder(roof_desc, sale_price_adj, .fun = median)) %>% 
  add_count(roof_desc) %>% 
  ggplot(aes(sale_price_adj, roof_desc, fill = n)) +
  geom_boxplot(outlier.alpha = 0,
               color = "grey") +
  scale_x_log10(label = dollar) +
  scale_fill_viridis_c()
```

#### Basement

Impute missing based on mode for `geo_id` and `style_desc.`
```{r}
assessments_valid %>% 
  mutate(basement_desc = fct_reorder(basement_desc, sale_price_adj, .fun = median)) %>% 
  add_count(basement_desc) %>% 
  ggplot(aes(sale_price_adj, basement_desc, fill = n)) +
  geom_boxplot(outlier.alpha = 0,
               color = "grey") +
  scale_x_log10(label = dollar) +
  scale_fill_viridis_c()
```

#### Basement garage

Impute missing based on mode for `geo_id` and `style_desc.`
Need to look up what this column is. Number of cars that can fit in the basement garage?
```{r}
assessments_valid %>% 
  select(sale_price_adj, basement_garage) %>% 
  mutate(basement_garage = as.character(basement_garage),
         basement_garage = fct_explicit_na(basement_garage),
         basement_garage = fct_reorder(basement_garage, sale_price_adj, .fun = median)) %>% 
  add_count(basement_garage) %>% 
  ggplot(aes(sale_price_adj, basement_garage, group = basement_garage, fill = n)) +
  geom_boxplot(outlier.alpha = 0,
               color = "grey") +
  scale_x_log10(label = dollar) +
  scale_fill_viridis_c()
```

#### Fireplaces

Positive relationship between number of fireplaces and sale price, but most houses have 1 or 2. Consider changing to lgl `fireplace_flag` column where it checks `fireplaces` >= 1.
```{r}
assessments_valid %>% 
  add_count(fireplaces) %>% 
  ggplot(aes(sale_price_adj, fireplaces, group = fireplaces, fill = n)) +
  geom_boxplot(outlier.alpha = 0,
               color = "grey") +
  scale_x_log10(label = dollar) +
  scale_fill_viridis_c()
```

### Identify UI cutoffs

```{r}
assessments_valid %>% 
  ggplot(aes(lot_area)) +
  geom_density() +
  geom_vline(xintercept = 40000) +
  scale_x_log10()
```

```{r}
assessments_valid %>% 
  ggplot(aes(finished_living_area)) +
  geom_density() +
  scale_x_log10() +
  geom_vline(xintercept = 10000)
```